---
title: "ebnm_point_laplace vs. ebayesthresh"
author: "Jason Willwerscheid"
date: "01/12/2021"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

I compare our implementation of `ebnm_point_laplace` to `EBayesThresh::ebayesthresh`, which does something very similar. There are a few differences, maybe most significant of which is that `ebayesthresh` constrains the scale of the Laplace component to be between $1/3$ and $25$. (I'm not sure why this is the case.) This constraint leads to sometimes large differences in the log likelihood of the optimal solutions, as seen below.

I compare across six possible scenarios:

* The "true" prior can be null, point-Laplace, or point-$t_5$
* Standard errors can be identical or different

Each point below represents one set of simulated data (one seed). The blue points are cases where our implementation does at least as well as `ebayesthresh` (faster or better log likelihood). Red points are cases where `ebayesthresh` does better. When the prior is non-null, our implementation is faster and obtains a better solution every time. In the null case, our implementation arrives at slightly better solutions and is faster for heteroskedastic data, but a bit slower for homoskedastic data. Our implementation is also more robust: for $n = 100000$, `ebayesthresh` errored out for 3 of the 23 seeds tested, while `ebnm_point_laplace` returned results for all seeds.

```{r code}
suppressMessages(library(tidyverse))
library(microbenchmark)
library(ebnm)
library(EbayesThresh)

one_sim <- function(n, prior_family, homosked, seed) {
  set.seed(seed)
  
  pi0 <- rbeta(1, shape1 = 10, shape2 = 2)
  scale <- rgamma(1, shape = 4, rate = 1)
  mu <- 0
  
  if (prior_family == "null") {
    theta <- mu
  } else if (prior_family == "pl") {
    theta <- mu + rexp(n, 1 / scale) * sample(c(-1, 1), n, replace = TRUE)
    theta[rbinom(n, size = 1, prob = pi0) == 1] <- mu
  } else {
    theta <- mu + scale * rt(n, df = 5)
    theta[rbinom(n, size = 1, prob = pi0) == 1] <- mu
  }
  
  if (homosked) {
    s <- 1
  } else {
    s <- sqrt(rexp(n))
  }
  
  x <- theta + rnorm(n, sd = s)
 
  return(list(x = x, s = s)) 
}

one_mb <- function(n, prior_family, homosked, mb_times, seed) {
  dat <- one_sim(n, prior_family, homosked, seed)
  
  mb_res <- microbenchmark(
    theirres <- ebayesthresh(x = dat$x, sdev = dat$s, a = NA, threshrule = "mean", 
                             universalthresh = FALSE, verbose = TRUE),
    ourres <- ebnm_point_laplace(x = dat$x, s = dat$s, optmethod = "nohess_nlm", 
                                 control = list(gradtol = sqrt(.Machine$double.eps), stepmax = 5)),
    times = mb_times
  )
  mb_res <- summary(mb_res)
  
  theirllik <- ebnm:::loglik_point_laplace(dat$x, dat$s, theirres$w, theirres$a, 0)
  ourllik <- ebnm:::loglik_point_laplace(dat$x, dat$s, 1 - ourres$fitted_g$pi[1], 
                                         1 / ourres$fitted_g$scale[2], 0)
  res <- list()
  res$llik_diff <- ourllik - theirllik
  res$mean_tdiff <- mb_res$mean[2] / mb_res$mean[1]
  res$median_tdiff <- mb_res$median[2] / mb_res$median[1]
  
  return(res)
}

one_n <- function(n, mb_times, seeds) {
  optmethods <- c("ebayesthresh", "ebnm_pl")
  
  all_res <- tibble()
  
  for (seed in seeds) {
    for (prior_family in c("null", "pl", "heavy")) {
      for (homosked in c(TRUE, FALSE)) {
        mb_res <- one_mb(n, prior_family, homosked, mb_times, seed)
        
        all_res <- all_res %>%
          bind_rows(tibble(seed = seed, n = n, prior_family = prior_family, 
                           homosked = homosked, llik_diff = mb_res$llik_diff, 
                           mean_tdiff = mb_res$mean_tdiff, median_tdiff = mb_res$median_tdiff))
      }
    }
  }
  
  return(all_res)
}

plot_one_n <- function(res) {
  df <- res %>%
    mutate(log2_tdiff = -log2(mean_tdiff)) %>%
    mutate(x = paste0(prior_family, ", ", ifelse(homosked, "homo", "het"))) %>%
    mutate(x = fct_relevel(x, "null, homo", "null, het", "pl, homo", "pl, het",
                           "heavy, homo", "heavy, het"))

  plot(ggplot(df, aes(x = x, y = log2_tdiff, col = log2_tdiff >= 0)) +
    geom_jitter(width = 0.3) +
    labs(x = NULL, y = NULL) +
    ggtitle("Difference in timing: log2(EBayesThresh / ebnm)"))
  
  plot(ggplot(df, aes(x = x, y = llik_diff, col = llik_diff >= 0)) +
    geom_jitter(width = 0.3) +
    labs(x = NULL, y = NULL) +
    ggtitle("Difference in log likelihood: ebnm - EBayesThresh"))
}
```

## $n = 1000$

```{r n1000}
res1000 <- one_n(1000, mb_times = 20L, seeds = 1:20)
plot_one_n(res1000)
```

## $n = 10000$

```{r n10000}
res10000 <- one_n(10000, mb_times = 10L, seeds = 11:30)
plot_one_n(res10000)
```

## $n = 100000$

```{r n100000}
# ebayesthresh fails for seeds 1, 5, and 12:
res100000 <- one_n(100000, mb_times = 5L, seeds = c(2:4, 6:11, 13:23))
plot_one_n(res100000)
```
