---
title: "Testing point-normal optimization methods (estimated mode)"
author: "Jason Willwerscheid"
date: "01/12/2021"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---


```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

I [again](optmethods_pn.html) compare optimization methods for solving the EBNM problem with point-normal priors but now I estimate the mode as well as estimating $\pi_0$ and $\sigma^2$. The "true" prior I use is:
$$ \theta_i \sim 0.9 \delta_\mu + 0.1N(\mu, 3^2) $$
I set $\mu = 1$ and I set all standard errors to 1.

```{r code}
library(microbenchmark)
library(ebnm)
library(ggplot2)

test_n <- function(n, mb_times = 100L, seed = 666) {
  set.seed(seed)
  # true g is 0.9 delta_mu + 0.1 N(mu, 3^2)
  mu <- 1
  theta <- rnorm(n, mean = mu, sd = 3)
  theta[rbinom(n, size = 1, prob = 0.9) == 1] <- mu
  s <- 1
  x <- theta + rnorm(n, sd = s)

  test_res <- microbenchmark(
    tres <- ebnm_point_normal(x, s, mode = "estimate", optmethod = "trust",
                              control = list(fterm = sqrt(.Machine$double.eps))),
    ebnm_point_normal(x, s, mode = "estimate", optmethod = "nograd_lbfgsb",
                      control = list(control = list(pgtol = sqrt(.Machine$double.eps)))),
    ores <- ebnm_point_normal(x, s, mode = "estimate", optmethod = "lbfgsb",
                        control = list(control = list(pgtol = sqrt(.Machine$double.eps)))),
    ebnm_point_normal(x, s, mode = "estimate", optmethod = "nohess_nlm", 
                      control = list(gradtol = sqrt(.Machine$double.eps))),
    ebnm_point_normal(x, s, mode = "estimate", optmethod = "nograd_nlm", 
                      control = list(gradtol = sqrt(.Machine$double.eps))),
    nres <- ebnm_point_normal(x, s, mode = "estimate", optmethod = "nlm", 
                        control = list(gradtol = sqrt(.Machine$double.eps))),
    times = mb_times
  )

  levels(test_res$expr) <- c("trust", "L-BFGS-B (no grad)", "L-BFGS-B", 
                          "nlm (no hess)", "nlm (no grad)", "nlm")
  
  llik_res <- c(tres$log_likelihood, ores$log_likelihood, nres$log_likelihood)
  names(llik_res) <- c("trust", "L-BFGS-B", "nlm")
  
  return(list(mb_res = test_res, llik_res = llik_res))
}
```

## n = 1000

Difference from maximum log likelihood among all methods:

```{r n1000}
res1000 <- test_n(1000)
res1000$llik_res - max(res1000$llik_res)
```

Timing results: 

```{r n1000t}
autoplot(res1000$mb_res)
```

## n = 10000

Difference from maximum log likelihood among all methods:

```{r n10000}
res10000 <- test_n(10000)
res10000$llik_res - max(res10000$llik_res)
```

Timing results: 

```{r n10000t}
autoplot(res10000$mb_res)
```

## n = 100000

Difference from maximum log likelihood among all methods:

```{r n100000}
res100000 <- test_n(100000)
res100000$llik_res - max(res100000$llik_res)
```

Timing results: 

```{r n100000t}
autoplot(res100000$mb_res)
```

