---
title: "A comparison of optimization methods for solving the parametric ebnm problem"
author: "Jason Willwerscheid"
date: "01/12/2021"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

I only consider optimization methods `nlm`, `nohess_nlm`, and `trust` here, since previous tests suggested that these were the only regularly competitive methods. I do a more careful comparison, with sixteen possible scenarios:

* The prior family can be point-normal or point-laplace
* The "true" prior can be an element of the prior family or null
* Standard errors can be identical or different
* The mode can be estimated or not

I use `nlm` as the baseline and show timing differences for `nohess_nlm` and `trust`. For example, `tdiff = 2` means that `nlm` is faster by a factor of 2. Thus blue squares indicate cases where `nlm` is faster and red squares indicate cases where a competing method is faster. I use `microbenchmark` with differently simulated data for each iteration, and I give comparisons using both mean and median timings.

```{r code}
suppressMessages(library(tidyverse))
library(microbenchmark)
library(ebnm)

one_test <- function(n, pn, null_case, homosked, est_mu, optmethod) {
  pi0 <- rbeta(1, shape1 = 10, shape2 = 2)
  scale <- rgamma(1, shape = 4, rate = 1)
  
  if (est_mu) {
    mu <- runif(1, -20, 20)
  } else {
    mu <- 0
  }
  
  if (null_case) {
    theta <- mu
  } else {
    if (pn) {
      theta <- mu + rnorm(n, sd = scale)
    } else {
      theta <- mu + rexp(n, 1 / scale) * sample(c(-1, 1), n, replace = TRUE)
    }
    theta[rbinom(n, size = 1, prob = pi0) == 1] <- mu
  }
  
  if (homosked) {
    s <- 1
  } else {
    s <- sqrt(rexp(n))
  }
  
  x <- theta + rnorm(n, sd = s)

  if (optmethod == "nlm") {
    res <- ebnm_point_normal(x, s, optmethod = "nlm", 
                             control = list(gradtol = sqrt(.Machine$double.eps)))
  } else if (optmethod == "nohess_nlm") {
    res <- ebnm_point_normal(x, s, optmethod = "nohess_nlm", 
                             control = list(gradtol = sqrt(.Machine$double.eps)))
  } else if (optmethod == "trust") {
    res <- ebnm_point_normal(x, s, optmethod = "trust",
                             control = list(fterm = sqrt(.Machine$double.eps)))
  } else {
    stop("Invalid optmethod.")
  }

  return(res)
}

one_mb <- function(n, pn, null_case, homosked, est_mu, mb_times) {
  res <- microbenchmark(
    one_test(n, pn, null_case, homosked, est_mu, optmethod = "nlm"),
    one_test(n, pn, null_case, homosked, est_mu, optmethod = "nohess_nlm"),
    one_test(n, pn, null_case, homosked, est_mu, optmethod = "trust"),
    times = mb_times
  )
  
  return(res)
}

one_n <- function(n, mb_times) {
  optmethods <- c("nlm", "nohess_nlm", "trust")
  
  all_res <- tibble()
  
  for (pn in c(TRUE, FALSE)) {
    for (null_case in c(TRUE, FALSE)) {
      for (homosked in c(TRUE, FALSE)) {
        for (est_mu in c(TRUE, FALSE)) {
          mb_res <- summary(one_mb(n, pn, null_case, homosked, est_mu, mb_times), 
                            unit = "s")
          
          means <- tibble(n = n, pn = pn, null_case = null_case, 
                          homosked = homosked, est_mu = est_mu, sumstat = "mean", 
                          optmethod = optmethods, t = mb_res$mean)
          medians <- tibble(n = n, pn = pn, null_case = null_case, 
                            homosked = homosked, est_mu = est_mu, sumstat = "median", 
                            optmethod = optmethods, t = mb_res$median)
          
          all_res <- all_res %>%
            bind_rows(means) %>%
            bind_rows(medians)
        }
      }
    }
  }
  
  all_res <- all_res %>%
    group_by(n, pn, null_case, homosked, est_mu, sumstat) %>%
    summarize(trust = max(ifelse(optmethod == "trust", t, 0)) /
                max(ifelse(optmethod == "nlm", t, 0)),
              nohess_nlm = max(ifelse(optmethod == "nohess_nlm", t, 0)) /
                max(ifelse(optmethod == "nlm", t, 0)),
              .groups = 'drop')
  
  all_res <- all_res %>%
    gather("comparison_method", "tdiff", trust, nohess_nlm)
  
  return(all_res)
}

plot_one_n <- function(res) {
  res <- res %>%
    mutate(x = factor(paste0(ifelse(pn, "PN", "PL"), ifelse(est_mu, " (est mu)", ""))),
           y = factor(paste0(ifelse(homosked, "Homo, ", "Hetero, "),
                             ifelse(null_case, "null case", "non-null"))),
           facet = paste0(comparison_method, " (", sumstat, " timing)")) %>%
    mutate(x = fct_relevel(x, "PN", "PL", "PN (est mu)", "PL (est mu)"),
           y = fct_relevel(y, "Hetero, null case", "Homo, null case", 
                           "Hetero, non-null", "Homo, non-null"))
  
  plt <- ggplot(res, aes(x = x, y = y)) + 
    geom_raster(aes(fill = tdiff)) +
    scale_fill_gradient2(midpoint = 1) +
    labs(x = NULL, y = NULL) +
    facet_wrap(~facet) +
    ggtitle("Timing differences vs. nlm")
  
  return(plt)
}
```

## $n = 1000$

```{r n1000}
plot_one_n(one_n(1000, mb_times = 100L))
```

## $n = 10000$

```{r n10000}
plot_one_n(one_n(10000, mb_times = 100L))
```

## $n = 100000$

```{r n100000}
plot_one_n(one_n(100000, mb_times = 100L))
```