---
title: "A comparison of optimization methods for solving the parametric ebnm problem"
author: "Jason Willwerscheid"
date: "01/12/2021"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

I only consider optimization methods `nlm`, `nohess_nlm`, and `trust` here, since previous tests suggested that these were the only regularly competitive methods. I do a more careful comparison, with sixteen possible scenarios:

* The prior family can be point-normal or point-laplace
* The "true" prior can be an element of the prior family or null
* Standard errors can be identical or different
* The mode can be estimated or not

I use `nlm` as the baseline and show timing differences for `nohess_nlm` and `trust`. For example, `log2_tdiff = 1` means that `nlm` is faster by a factor of 2. Thus blue squares indicate cases where `nlm` is faster and red squares indicate cases where a competing method is faster. 

```{r code}
suppressMessages(library(tidyverse))
library(microbenchmark)
library(ebnm)

one_sim <- function(n, pn, null_case, homosked, est_mu, seed) {
  set.seed(seed)
  
  pi0 <- rbeta(1, shape1 = 10, shape2 = 2)
  scale <- rgamma(1, shape = 4, rate = 1)
  
  if (est_mu) {
    mu <- runif(1, -20, 20)
  } else {
    mu <- 0
  }
  
  if (null_case) {
    theta <- mu
  } else {
    if (pn) {
      theta <- mu + rnorm(n, sd = scale)
    } else {
      theta <- mu + rexp(n, 1 / scale) * sample(c(-1, 1), n, replace = TRUE)
    }
    theta[rbinom(n, size = 1, prob = pi0) == 1] <- mu
  }
  
  if (homosked) {
    s <- 1
  } else {
    s <- sqrt(rexp(n))
  }
  
  x <- theta + rnorm(n, sd = s)
  
  return(list(x = x, s = s))
}

one_mb <- function(n, pn, null_case, homosked, est_mu, mb_times, seed) {
  dat <- one_sim(n, pn, null_case, homosked, est_mu, seed)
  x <- dat$x
  s <- dat$s

  if (pn) {
    ebnm_fn <- ebnm_point_normal
  } else {
    ebnm_fn <- ebnm_point_laplace
  }
  
  if (est_mu) {
    ebnm_mode <- "estimate"
  } else {
    ebnm_mode <- 0
  }
  
  mb_res <- microbenchmark(
    nlm_res <- do.call(ebnm_fn, 
                       list(x = x, s = s, mode = ebnm_mode, optmethod = "nlm", 
                            control = list(gradtol = sqrt(.Machine$double.eps)))),
    nohess_res <- do.call(ebnm_fn, 
                          list(x = x, s = s, mode = ebnm_mode, optmethod = "nohess_nlm", 
                               control = list(gradtol = sqrt(.Machine$double.eps)))),
    trust_res <- do.call(ebnm_fn, 
                         list(x = x, s = s, mode = ebnm_mode, optmethod = "trust",
                              control = list(fterm = sqrt(.Machine$double.eps)))),
    times = mb_times
  )
  mb_res <- summary(mb_res)
  
  res <- list()
  res$nohess_mean_tdiff <- mb_res$mean[2] / mb_res$mean[1]
  res$nohess_median_tdiff <- mb_res$median[2] / mb_res$median[1]
  res$trust_mean_tdiff <- mb_res$mean[3] / mb_res$mean[1]
  res$trust_median_tdiff <- mb_res$median[3] / mb_res$median[1]
  
  return(res)
}

one_n <- function(n, mb_times, seeds) {
  all_res <- tibble()
  
  for (seed in seeds) {
    for (pn in c(TRUE, FALSE)) {
      for (null_case in c(TRUE, FALSE)) {
        for (homosked in c(TRUE, FALSE)) {
          for (est_mu in c(TRUE, FALSE)) {
            mb_res <- one_mb(n, pn, null_case, homosked, est_mu, mb_times, seed)
            
            all_res <- all_res %>%
              bind_rows(tibble(seed = seed, n = n, pn = pn, null_case = null_case, 
                               homosked = homosked, est_mu = est_mu,
                               optmethod = "nohess", 
                               mean_tdiff = mb_res$nohess_mean_tdiff, 
                               median_tdiff = mb_res$nohess_median_tdiff)) %>%
              bind_rows(tibble(seed = seed, n = n, pn = pn, null_case = null_case, 
                               homosked = homosked, est_mu = est_mu,
                               optmethod = "trust", 
                               mean_tdiff = mb_res$trust_mean_tdiff, 
                               median_tdiff = mb_res$trust_median_tdiff))
          }
        }
      }
    }
  }

  return(all_res)
}

plot_one_n <- function(res) {
  df <- res %>%
    group_by(n, pn, null_case, homosked, est_mu, optmethod) %>%
    summarize(log2_tdiff = log2(mean(mean_tdiff)), .groups = 'drop') %>%
    mutate(x = factor(paste0(ifelse(pn, "PN", "PL"), ifelse(est_mu, " (est mu)", ""))),
           y = factor(paste0(ifelse(homosked, "Homo, ", "Hetero, "),
                             ifelse(null_case, "null case", "non-null")))) %>%
    mutate(x = fct_relevel(x, "PN", "PN (est mu)", "PL", "PL (est mu)"),
           y = fct_relevel(y, "Hetero, null case", "Homo, null case", 
                           "Hetero, non-null", "Homo, non-null"))
  
  plt <- ggplot(df, aes(x = x, y = y)) + 
    geom_raster(aes(fill = log2_tdiff)) +
    scale_fill_gradient2(midpoint = 0) +
    labs(x = NULL, y = NULL) +
    facet_wrap(~optmethod) +
    ggtitle("Timing differences vs. nlm")
  
  return(plt)
}

# one_plot <- function(res, plot_pn, plot_est_mu) {
#   df <- res %>%
#     filter(pn == plot_pn, est_mu == plot_est_mu)
#   
#   df <- df %>%
#     mutate(log2_tdiff = log2(mean_tdiff)) %>%
#     mutate(x = paste0(ifelse(null_case, "null", "nonnull"), ", ", 
#                       ifelse(homosked, "homo", "het"))) %>%
#     mutate(x = fct_relevel(x, "null, homo", "null, het", "nonnull, homo", "nonnull, het"))
# 
#   plot(ggplot(df, aes(x = x, y = log2_tdiff)) +
#          geom_boxplot() + 
#          geom_jitter(width = 0.3) +
#          labs(x = NULL, y = NULL) +
#          geom_hline(aes(yintercept = 0), linetype = "dashed") + 
#          ggtitle(paste0("log2 difference in timing vs. nlm: ",
#                         ifelse(plot_pn, "point-normal", "point-Laplace"), ", ",
#                         ifelse(plot_est_mu, "estimate mu", "fix mu"))) +
#          facet_wrap(~optmethod))
# }
# 
# plot_one_n <- function(res) {
#   for (pn in c(TRUE, FALSE)) {
#     for (est_mu in c(FALSE, TRUE)) {
#       one_plot(res, pn, est_mu)
#     }
#   }
# }
```

## $n = 1000$

```{r n1000}
res1000 <- one_n(1000, mb_times = 20L, seeds = 1:20)
plot_one_n(res1000)
```

## $n = 10000$

```{r n10000}
res10000 <- one_n(10000, mb_times = 10L, seeds = 1:20)
plot_one_n(res10000)
```

## $n = 100000$

```{r n100000}
res100000 <- one_n(100000, mb_times = 5L, seeds = 1:20)
plot_one_n(res100000)
```
